--- 
title: "Medizinische Informatik: Zusatzweiterbildung"
author: "Grigorij Schleifer"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "Meine Notizen für die Prüfung"
---

# Preface {-}

Ich bin kein Autor und schreibe viel zu selten um mich als jemanden zu sehen der gut schreibt. Ich bin Anästhesist und Notfallmediziner und seit einigen Jahren bin ich interessiert an medizinischer Informatik mit v.a. Bezug zu Bearbeitung, Specherung und einfach Umgang mit medizinischen Daten. Da ich das Glück haben auch einen Teil meiner klinischen Beschäftigung in Richtung Medizinischer Informatik zu verlagern protokolliere ich hier meinen Lernweg und sammle gleichzeitig meine Gedanken zu den Weiterbildungsinhalten "Medizinische Informatik". Es wäre natürlich total vermessen die Inhalte für die Weiterbildung selbst zusammenzustellen und ich haben im Moment auch nicht die leiseste Ahnung was alles so medizinisch Informatik ist aber erfreulicherweise gibt es genug Ressourcen von unseren Ärztekammern die mir die Arbeit sehr erleichtern. Dieses Buch habe ich also nach den Vorgaben der Weiterbildungsordnung 2018 der Bundesärztekammer gegliedert und die vorgeschlagenen Module als Kapiteln übernommen. Alle links sind in den entschprechenden Kapiteln enthalten.  

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

<!--chapter:end:index.Rmd-->

# Angewandte Informatik {#intro}


You can label chapter and section titles using `{#label}` after them, e.g., we can reference Chapter \@ref(literature). If you do not manually label them, there will be automatic labels anyway, e.g., Chapter \@ref(05-Statistik).

Figures and tables with captions will be placed in `figure` and `table` environments, respectively.

```{r nice-fig, fig.cap='Here is a nice figure!', out.width='80%', fig.asp=.75, fig.align='center'}
par(mar = c(4, 4, .1, .1))
plot(pressure, type = 'b', pch = 19)
```

Reference a figure by its code chunk label with the `fig:` prefix, e.g., see Figure \@ref(fig:nice-fig). Similarly, you can reference tables generated from `knitr::kable()`, e.g., see Table \@ref(tab:nice-tab).

```{r nice-tab, tidy=FALSE}
knitr::kable(
  head(iris, 20), caption = 'Here is a nice table!',
  booktabs = TRUE
)
```

You can write citations, too. For example, we are using the **bookdown** package [@R-bookdown] in this sample book, which was built on top of R Markdown and **knitr** [@xie2015].

<!--chapter:end:01-Modul-I.Rmd-->

# Datenschutz

Here is a review of existing methods.

<!--chapter:end:02-Modull-II.Rmd-->

# Medizinische Statistik

```{r nice-tab, tidy=FALSE}
Inhaltsverzeichnis
4.1 Modul I – Angewandte Informatik
4.2 Modul II – Datenschutz/Datensicherheit 
4.3 Modul III – Medizinische Statistik/Epidemiologie/Biometrie (15 
4.4 Modul IV – Medizinische Dokumentation (40 h) 
4.5 Modul V – Bildverarbeitung/Biosignalverarbeitung (15 
4.6 Modul VI – Entscheidungsunterstützung (20 
4.7 Modul VII – Informations- und Kommunikationssysteme (40 
4.8 Modul VIII – Management in der Gesundheits-IT (15 
4.9 Modul IX – Informationsmanagement/Klinisches Datenmanagement/ Klinische Forschung (30 h) .

4.10 Modul X – Telemedizin und Telematik (15 
```



Abkürzung Bedeutung
AAL Ambient Assisted Living
AMG Arzneimittelgesetz
AMTS Arzneimitteltherapiesicherheit
ATC-Klassifikation Anatomisch-Therapeutisch-Chemische Klassifikation
BDSG Bundesdatenschutzgesetz
BPMN Business Process Model and Notation
BSI Bundesamt für Sicherheit in der Informationstechnik
CPOE Computerized Physician Order Entry
CRM Customer-Relationship-Management
DICOM Digital Imaging and Communications in Medicine
DRG Diagnosis related groups
DSGVO Datenschutz-Grundverordnung
EBM Einheitlicher Bewertungsmaßstab
ER-Modell Entity-Relationship-Modell
GMDN Global Medical Device Nomenclature
GOÄ Gebührenordnung für Ärzte
HL7 Health Level 7
HTA Health Technology Assessment
HTML Hypertext Markup Language
ICD-10 International Classification of Diseases, Rev. 10
ICD-O International Classification of Diseases for Oncology
ICF International Classification of Functioning, Disability and Health
IHE Integrating the Healthcare Enterprise
ITIL Information Technology Infrastructure Library
JSON JavaScript Object Notation
KIS Krankenhausinformationssystem
KISS Krankenhaus-Infektions-Surveillance-System
LOINC Logical Observation Identifiers Names and Codes
MPG Medizinproduktegesetz
OH KIS Orientierungshilfe Krankenhausinformationssysteme
OPS Operationen- und Prozedurenschlüssel
PACS Picture Archiving and Communication System
PDMS Patientendatenmanagementsystem
PEPP Pauschalierendes Entgeltsystem Psychiatrie und Psychosomatik
PLM Product-Lifecycle-Management
SCM Software-Configuration-Management
SNOMED-CT Systematized Nomenclature of Human and Veterinary Medicine – Clinical Terms
SOAP Simple Object Access Protocol
SPSS Statistical Package for the Social Sciences
SQL Structured Query Language
TNM-Klassifikation Tumour-Node-Metastasis-Klassifikation
TOM Technische und organisatorische Maßnahmen
UCUM Unified Code for Units of Measure
UML Unified Modeling Language
XML Extensible Markup Language

## Methoden Nummer 2
We describe our methods in this chapter.

Math can be added in body using usual syntax like this 

## math example

$p$ is unknown but expected to be around 1/3. Standard error will be approximated

$$
SE = \sqrt{\frac{p(1-p)}{n}} \approx \sqrt{\frac{1/3 (1 - 1/3)} {300}} = 0.027
$$

You can also use math in footnotes like this^[where we mention $p = \frac{a}{b}$].

We will approximate standard error to 0.027[^longnote]

[^longnote]: $p$ is unknown but expected to be around 1/3. Standard error will be approximated

    $$
    SE = \sqrt{\frac{p(1-p)}{n}} \approx \sqrt{\frac{1/3 (1 - 1/3)} {300}} = 0.027
    $$

<!--chapter:end:03-Modul-III.Rmd-->

# Medizinische Dokumentation

Some _significant_ applications are demonstrated in this chapter.

## Example one

## Example two

<!--chapter:end:04-Modul-IV.Rmd-->

# Bildverarbeitung

```{r}
# https://builtin.com/data-science/boxplot
# Median (Q2/50th percentile): The middle value of the data set.
# First Quartile (Q1/25th percentile): The middle number between the smallest number (not the “minimum”) and the median of the data set.
# Third Quartile (Q3/75th percentile): The middle value between the median and the highest value (not the “maximum”) of the dataset.
# Interquartile Range (IQR): 25th to the 75th percentile.
# Whiskers (shown in blue)
# Outliers (shown as green circles)
# “Minimum”: Q1 - 1.5*IQR
# “Maximum”: Q3 + 1.5*IQR
```


<!--chapter:end:05-Modul-V.Rmd-->

# Entscheidungsunterstützung

<!--chapter:end:06-Modul-VI.Rmd-->

# Informationssysteme 

<!--chapter:end:07-Modul-VII.Rmd-->

# Management

<!--chapter:end:08-Modul-VIII.Rmd-->

# Informationsmanagement
## Klinisches Datenmanagement
## Klinische Forschung

<!--chapter:end:09-Modul-IX.Rmd-->

# Telemedizin
## Telematik

<!--chapter:end:10-Modul-X.Rmd-->

